High geomagnetic field intensity recorded by anorthosite xenoliths requires a strongly powered late Mesoproterozoic geodynamo

### 2022-01-28 PNAS_manuscript/PNAS_manuscript.tex

The lack of symplecites and the lack of large plagioclase-hosted titanomagnetite needles within the Beaver River anorthosite xenoliths targetted in this study distinguish them from plagioclase cumulates of the Duluth Complex and other layered mafic complexes where anisotropic magnetic mineral fabrics associated with igneous foliation often occur \cite{Scofield1986a, Selkin2000a, Feinberg2006a}. The relative lack of fabric and associated preferred orientation makes the Beaver River anorthosite xenoliths a particularly compelling target for paleointensity experiments.

@@

In addition to estimating paleointensity values by introducing a set of selection criteria to filter our experiment results, we apply an independent method of \citealp{Cych2021a} which uses all experimental data regardless of their NRM/TRM plot statistics to perform bias corrected estimation of paleointensity (BiCEP). This Bayesian probabilistic method is based on an assumption that paleointensity estimates from specimens that come from a same cooling unit are distributed around a true paleointensity value with the various deflections being expressed as the curvature parameter of the NRM/TRM plot \cite{Paterson2011a}. The posterior paleointensity distributions from these sites with high-quality specimen-level data are in agreement with the site-level averages developed using the selection criteria approach (Fig. \ref{fig:PINT_cooling_corrected}; Fig. S1). Overall, the high-quality results from the anorthosite xenoliths of the Beaver River diabase indiacte that the anorthosite xenoliths record a high geomagnetic field ca. 1092 Ma.

@@

The hypothesis that a weak Ediacaran geomagnetic field is a telltale sign of the lack of an inner core with core nucleation following shortly thereafter may predict that it is the most significant weak to strong field transition in the paleointensity record. However, Fig. \ref{fig:PINT_compilation} shows that transitions from low to high field intensities occurred before, during, and after the Ediacaran. In the Ediacaran record developed to date, there is a two-fold increase in Earth's virtual dipole moment when comparing estimates from the ca. 565 Ma Sept-$\hat{I}$les intrusions \cite{Bono2019a} to those from ca. 550 Ma volcanics of the Skinner Cove Formation \cite{Thallner2021a} (Fig. \ref{fig:PINT_compilation}). In the late Mesoproterozoic, there is at least a six-fold increase within a period of $\sim$35 Myr from a low average virtual dipole moment of $\sim$13 ZAm$^2$ recorded by the ca. 1140 Ma Abitibi dikes which yielded straight Arai plots \cite{Macouin2003a}, to a high moment of $\sim$70 ZAm$^2$ recorded by the ca. 1106 Ma Osler Volcanics, with even stronger values from ca. 1092 Ma by Beaver River anorthosite xenoliths that record virtual dipole moments up to $\sim$140 ZAm$^2$. While the Abitibi dikes paleointensity estimates do not go to values as low as the Sept-$\hat{I}$les intrusions, this virtual dipole moment increase in the Mesoproterozoic is the largest yet documented in the Precambrian on a 10s of millions of year timescale (Fig. \ref{fig:PINT_compilation}). The tempo and scale of this field intensity transition could match with model-based predictions associated with the onset of inner core nucleation \cite{Davies2021a}. This timing would be broadly consistent with that proposed by \citealp{Biggin2015a}. However, a model prediction could be of sustained strong field values following inner core nucleation which is challenged by data from the ca. 1070 Ma Bangemall Sills which include a sill with a low virtual dipole moment of $\sim$6.4 ZAm$^2$ (Fig. \ref{fig:PINT_compilation};  \citealp{Lloyd2021b}). Following the Ediacaran, there are also low paleointensity estimates from Devonian rocks such as ca. 370 Ma dikes and lavas of the Siberian Viluy Traps that give virtual dipole moment of 4.3 to 14.9 ZAm$^2$ (Fig. \ref{fig:PINT_compilation}; \citealp{Hawkins2019a}). These low values and as well as data from the ca. 414 Ma Strathmore lava flows \cite{Hawkins2021a}, the ca. 410-380 Ma lava flows of the Minusa Basin and the Kola Peninsula \cite{Shcherbakova2017a}, the ca. 408–393 Ma Buribay volcanics  volcanic complex \cite{Shcherbakova2021a} and the ca. 332 Ma Kinghorn volcanics \cite{Hawkins2021a} has led to the proposal of this interval as the ``Mid-Paleozoic Dipole Low'' \cite{Hawkins2021a}. This  ``Mid-Paleozoic Dipole Low'' is followed by high paleointensity values such that there is a six-fold increase in the database from virtual dipole moment of $\sim$16 ZAm$^2$ at ca. 332 Ma to $\sim$99 ZAm$^2$ ca. 308 Ma in the late Carboniferous \cite{Hawkins2021a}.

Given that each record of a weak field can not be a minimum prior to inner core nucleation, what processes could lead to a weak dipole at Earth's surface even in the presence of a crystallizing inner core? Numerical models have shown that the dipole moment is sensitive to both the magnitude and spatial pattern of heat flow across the core-mantle boundary when there are strong available power sources to the geodynamo \cite{Olson2007a, Olson2010a}. In such models, relatively low total heat flux across the core-mantle boundary can prevent the axial dipole from reversing whereas a high heat flux through the boundary can result in an increase in reversal frequency and decrease in the axial dipole intensity. The ``Mid-Paleozoic Dipole Low'' has been hypothesized to be the result of elevated core-mantle boundary heat flux condition at a time when there was also available power from a crystallizing inner core \cite{Hawkins2019a}, thereby explaining the observed low paleointensities which include values as weak as those of the ca. 565 Ma Sept-$\hat{I}$les intrusions \cite{Bono2019a} (Fig. \ref{fig:PINT_compilation}). Mantle convection can modulate core mantle boundary heat flow through changes in the structure of the deep mantle associated with upwelling plumes \cite{Larson1991a, Courtillot2007a} and subducted slabs \cite{Tan2002b, Biggin2012a, Hounslow2018a}. Strong evidence for differential plate tectonic motion extends back to ca. 2.2 Ga in the Paleoproterozoic \cite{Mitchell2014a,Swanson-Hysell2021b} and potentially back to ca. 3.2 Ga in the Archean \cite{Brenner2020a}. Plate tectonic modulations of core mantle boundary heat flow are therefore expected throughout the Proterozoic. Such changes may explain large variability in Proterozoic paleointensity values similar to those seen in the Phanerozoic \cite{Lloyd2021a} and may challenge our ability to detect the increase in surface geomagnetic field strength predicted to have happened at the onset of inner core crystallization.

Overall, the high-fidelity paleointensity recorders of the Beaver River anorthosite xenoliths in the well-preserved Midcontinent Rift record strong field strengths 1.1 billion years ago. The highest site level field strength would rank in the 99 percentile of the paleointensity database for the Cenozoic Era (the last 66 million years) when there was unequivocally a crystallizing inner core. These high surface field strengths necessitate appreciable power to the late Mesoproterozoic geodynamo.

% Could core-mantle interactions similar to that today exist before the nucleation of the inner core and modulated the geodynamo? With emerging paleomagnetic evidence supporting the existence of modern-day-like plate tectonics that extends through the Proterozoic \cite{Swanson-Hysell2021c} and to the Archean \cite{Mitchell2014a}, kinetic and thermal energy exchange between deep and shallow Earth likely existed since the early time of Earth's history. If the mantle forcing associated with these activities exists regardless of the present of inner core, it could have contributed to the anomalously low or high paleointensity values observed during the Phanerozoic and Proterozoic (Fig. \ref{fig:PINT_compilation}).

% heterogeneity through the lower mantle can weaken and destabilize the geomagnetic axial dipole, reduce its moment, and increase reversal frequency. with non-reversing behavior at low heat flow and frequent reversals at high heat flow conditions.
% We need to say models indicate that there can be weak fields when there are strong available power sources to the geodynamo. Cite Olson 2007 and Olson et al. 2010. The ``Mid-Paleozoic Dipole Low'' has been interpreted to be the result of high total core-mantle boundary heat flux thereby explaining field nearly as weak as those of the Sept-Iles data in a framework that includes available power from a crystallizing inner core (can cite Hawkins et al. 2019 as well as other Biggin papers that make this point). There can also be large changes in the strength of the dipole due to changes in the geometry of core-mantle boundary heat flux (these can be connected to subducted slabs and upwelling).

% An alternative hypothesis for the multiple transitions from low to high field intensities in the middle Paleozoic and the Proterozoic is that they are manifestations of secularly changing mantle forcing of the geodynamo across the core–mantle boundary which may be unrelated to the growth of the inner core. The time scale over which these changes happen may be on the order of less than one million years. During the Pleistocene, when we known a present-day-like geodynamo existed, paleointensity records show at least a six-fold fluctuation of the geomagnetic dipole moment can happen (Fig. \ref{fig:PINT_compilation}). 
%

%The low paleointensity records from the ca. 720 Ma dolerite dikes of the rapidly emplaced Franklin Large Igneous Province could be one example of capturing these short-term variations and not fully averaging the secular changes of the geomagnetic field (Fig. \ref{fig:PINT_compilation}; \citealp{Lloyd2021a, Lloyd2021b}).

% In the Neoproterozoic, ca. 720 Ma dolerite dikes of the Franklin Large Igneous Province record a mean virtual dipole moment of  In the Phanerozoic, multiple low paleointensities has been observed in Devonian-aged volcanic units (Fig. \ref{fig:PINT_compilation}; \citealp{Shcherbakova2017a, Shcherbakova2021a, Hawkins2021a}).

%The observed paleointensity variations from the Midcontinent Rift also question the attribution to the onset of inner core nucleation as the unique cause for driving a significant paleointensity increase.  Besides the

% Accompanying the low paleointensity values, paleodirectional data from the Ediacaran-Cambrian \cite{Bono2015a, Halls2015b, Kodama2021a} and the Devonian show anomalous directional fluctuations \cite{Shcherbakova2019a} which could be consistent with there being a largely diminished dipole component of Earth's magnetic field. If the weak paleointensities in the Ediacaran or the Devonian are attributed to the geodynamo being on the brink of collapse just prior to the nucleation of the solid inner core.

% if one invokes inner core nucleation for all these low to high transitions, the data would suggest the inner core nucleated multiple times. Alternatively, it could that all except one are not related to the inner core nucleation, but merely a manifestation of short term dynamo variations. 
% n narrow time scales regardless of the existence of a solid inner core. Therefore, low paleointensity values or short-term fluctuations may not be sufficient evidence for interpreting Earth's inner core status through history.

% The high geomagnetic field estimates from the Midcontinent Rift volcanics and intrusives are not the only suite of high paleointensity records at odds with the hypothesis of a monotonically decaying geodynamo through the entire Proterozoic Eon \cite{Biggin2009a, Bono2019a}. The $\sim$60 ZAm$^2$ virtual dipole moment estimates from the ca. 755 Ma Mundine Well dikes are much higher than the values predicted by the monotonic decrease model of \cite{Bono2019a}. Although one may consider these data capturing relatively short term fluctuations of the geomagnetic field, they nevertheless indicate that the geodynamo had strong power sources during the Proterozoic. On the other hand, besides the low paleointensity values recorded by the ca. 1140 Ma Abitibi dikes, low estimates have been obtained from the ca. 1070 Ma Bangemall Sills \cite{Lloyd2021b}, the ca. 720 Ma dolerite dikes of the Franklin Large Igneous Province \cite{Lloyd2021a}, as well as Devonian-aged rocks from the Minusa Basin, Kola Peninsula, Buribay volcanic complex and volcanic rocks from the East European Craton (Fig. \ref{fig:PINT_compilation};  \citealp{Shcherbakova2017a, Shcherbakova2019a, Shcherbakova2021a}). The multiple occurrences of low geomagnetic field intensities during the Proterozoic and the Paleozoic indicate that a waning thermal convection regime in Earth's before the emergence of the solid inner core may not be the solely possible mechanism for triggering an weak dipole field.

% In the meantime, paleomagnetic directional anomalies may be another suite of intriguing observational records that is worth investigating. The anomalous directional fluctuations during the Ediacaran and Devonian \cite{Bono2015a, Kodama2021a, Shcherbakova2019a} perhaps can be consistent with there being a weak geodynamo regime when dipole component of Earth's field was largely diminished.

% conclude on whether an apparent rise or decay within a narrow time period ($\sim$35 Ma) during the Proterozoic is associated with transitions in the core phase transition or is a result of short-term excursion of the geomagnetic field that may be common through much of the evolution of Earth's core.

% an apparent rapid decrease in geomagnetic dipole moment from ca. 1092 Ma to ca. 1070 Ma can be evidenced by the contrasting high values from the Midcontinent Rift anorthosite xenoliths and the low values from the Bangemall Sills (Fig. \ref{fig:PINT_compilation};  \citealp{Lloyd2021b}). These data suggest that an apparent transition of virtual dipole moment similar in scale with what is predicted for a dynamo powered with or without compositional convection in the core may not be unique during the Precambrian. In addition, given that a five-fold fluctuation of the geomagnetic dipole moment exist in the paleointensity records for the past 10 million years, when we know a modern-day-like geodynamo likely powered by both thermal and compositional convection existed, we cannot conclude on whether an apparent rise or decay within a narrow time period ($\sim$35 Ma) during the Proterozoic is associated with transitions in the core phase transition or is a result of short-term excursion of the geomagnetic field that may be common through much of the evolution of Earth's core. Recently, ultra-low paleointensities (2.01–7.07) $\mu$T from the Devonian has been found by  \cite{Shcherbakova2021a}. These data suggest that significant excursion of low geomagnetic field intensities may occur regardless of the inner core formation.

% These data suggest that an apparent transition of virtual dipole moment similar in scale with what is predicted for a dynamo powered with or without compositional convection in the core may not be unique during the Precambrian. In addition, given that a five-fold fluctuation of the geomagnetic dipole moment exist in the paleointensity records for the past 10 million years, when we know a modern-day-like geodynamo likely powered by both thermal and compositional convection existed, we cannot conclude on whether an apparent rise or decay within a narrow time period ($\sim$35 Ma) during the Proterozoic is associated with transitions in the core phase transition or is a result of short-term excursion of the geomagnetic field that may be common through much of the evolution of Earth's core.

% Paleomagnetic observational records near the Ediacaran-Cambrian boundary have been interpreted to be consistent with a trend of decreasing axial dipole strength toward a very low value \cite{Lloyd2021a, Bono2019a}, followed by a rapid recovery (a few tens of million years) toward higher values soon after \cite{Thallner2021a}. Such a decrease-increase transition can be consistent with dynamo numerical simulations which adopt a rather high thermal conductivity of the core and projects an inner core birth near this time \cite{Gomi2013a, Davies2015a, Labrosse2015a, Ohta2016a}. In particular, some models have suggested that the kick-start of rigorous compositional convection between the inner core and the outer core due to the onset of inner core nucleation can result in almost an order of magnitude increase in Earth's axial dipole moment \cite[e.g.][]{Davies2021a}. The enriching observational records from the Ediacaran and Cambrian have been suggested to indicate that a geodynamo driven solely by thermal convection was waning to its brink of collapse near the end of Ediacaran \cite{Bono2019a, Thallner2021b} and was recovering toward modern-day strengths in the Cambrian \cite{Thallner2021a}.

% However, the updated paleointensity records from the Midcontinent Rift together with previous results from other Mesoproterozoic rocks suggest that the a dramatic increase in geomagnetic dipole moment similar in scale to that modeled for after the onset of inner core nucleation may not be unique to the Ediacaran-Cambrian boundary (Fig. \ref{fig:PINT_compilation}). The ca. 1140 Ma Abitibi dikes \cite{Macouin2003a} record an average virtual dipole moment of $\sim$13 ZAm$^2$, similar to the low values of $\sim$7 ZAm$^2$ obtained from the single silicate crystals of the ca. 565 Ma Sept-$\hat{I}$les intrusions \cite{Bono2019a}. However, the relatively high moments of $\sim$72 ZAm$^2$ from the ca. 1105 Ma Osler Volcanic Group \cite{Sprain2018a} and $\sim$89 ZAm$^2$ from the ca. 1092 Ma Beaver River anorthosite xenoliths (Fig. \ref{fig:PINT_compilation}) are about six times the values from the Abitibi dikes. This increase in absolute dipole moment estimates during $\sim$50 Myr recorded by the Midcontinent Rift rocks are much higher than that in the Ediacaran (Fig. \ref{fig:PINT_compilation}).

% observation of an increase or a decrease in virtual dipole moment may not be unique during the Precambrian. A sparse spatial and temporal coverage of the Precambrian paleointensity records may not be sufficient for further analyses of the long-term evolution of the geodynamo. Given that a five-fold fluctuation of the geomagnetic dipole moment exist in the paleointensity records for the past 10 million years (5 and 95 percentiles), when we know a modern-day-like geodynamo likely powered by both thermal and compositional convection existed, we cannot conclude on whether an apparent rise or decay within a narrow time period ($\sim$35 Ma) during the Proterozoic is associated with transitions in the core fluid regime or is a result of short-term variations of the geomagnetic field that may be common through much of the evolution of Earth's core. Regardless, the high geomagnetic dipole field moment estimate from the Beaver River anorthosite xenoliths necessitates that there being a vigorous late Mesoproterozoic geodynamo.

% Overall, the existing absolute paleointensity records in the PINT database remains sparse in terms of spatial and temporal coverage as compared to more geologically recent times. (Fig. \ref{fig:PINT_compilation}). More data are needed in order to better characterize the evolution of Earth's dynamo and thermal evolution.

% \section*{Conclusion}

% The paucity of Precambrian paleointensity estimates, as well as non-ideal results within the existing database, present challenges when seeking to interpret long-term trends in the strength of the geomagnetic field. The Midcontinent Rift of North America presents an opportunity to develop robust paleointensity estimates spanning a $\sim$25 Myr interval of the late Mesoproterozoic. Rock magnetic data support that the ca. 1092 Ma anorthosite xenoliths of the Midcontinent Rift diabase can have dominantly high coercivity magnetic mineralogy with minimal anisotropic or nonlinear thermal remanence acquisition, producing high quality paleointensity results. Their overall high estimates of the late Mesoproterozoic virtual dipole moment together with previous Midcontinent Rift records suggest that a strong geomagnetic field like today's might have persisted through the rifting.

% The new data in context of the current Precambrian paleointensity database suggest that an observation of a strong increase in geomagnetic dipole moment may not be unique during the Precambrian. More spatial and temporal coverage of the observational records is needed for further analyses of the long-term evolution of the geodynamo.

@@

% After the paleointensity experiment were finished, the same specimens underwent additional TRM anisotropy checks with full TRMs given at 0, 30, 60, 90, 120, 150 degrees in the horizontal plane in specimen coordinates. After that, full TRM acquisition experiments were performed on the same specimens with applied fields of 30, 50, 70, and 90 $\mu$T along specimen vertical axis.

@@

% \begin{table*}
% \caption{Quality criteria used for selecting paleointensity data. See the text for more detail.} 
% \centering
% \begin{tabular}{ccccccccc}
% \hline
% MAD (\textdegree) & Beta (per cent) & DANG (\textdegree) & FRAC& SCAT & GAP-MAX & NpTRM & N Arai\\
% \hline
% 20 & 15 & 5 & 0.6 & TRUE & 0.25 & 2 & 4 \\
% \hline

% \end{tabular}
% \label{tab:criteria}
% \end{table*}

@@

% \subsection*{Manuscript Length}

% A standard 6-page article is approximately 4,000 words, 50 references, and 4 medium-size graphical elements (i.e., figures and tables). The preferred length of articles remains at 6 pages, but PNAS will allow articles up to a maximum of 12 pages.

% \subsection*{References}

% References should be cited in numerical order as they appear in text; this will be done automatically via bibtex, e.g. \cite{belkin2002using} and \cite{berard1994embedding,coifman2005geometric}. All references cited in the main text should be included in the main manuscript file.

% \subsection*{Data Archival}

% PNAS must be able to archive the data essential to a published article. Where such archiving is not possible, deposition of data in public databases, such as GenBank, ArrayExpress, Protein Data Bank, Unidata, and others outlined in the \href{https://www.pnas.org/page/authors/journal-policies#xi}{Information for Authors}, is acceptable.

% \begin{SCfigure*}[\sidecaptionrelwidth][t]
% \centering
% \includegraphics[width=11.4cm,height=11.4cm]{frog}
% \caption{This legend would be placed at the side of the figure, rather than below it.}
% \label{fig:side}
% \end{SCfigure*}

% \subsection*{Digital Figures}

% EPS, high-resolution PDF, and PowerPoint are preferred formats for figures that will be used in the main manuscript. Authors may submit PRC or U3D files for 3D images; these must be accompanied by 2D representations in TIFF, EPS, or high-resolution PDF format. Color images must be in RGB (red, green, blue) mode. Include the font files for any text.

% Images must be provided at final size, preferably 1 column width (8.7cm). Figures wider than 1 column should be sized to 11.4cm or 17.8cm wide. Numbers, letters, and symbols should be no smaller than 6 points (2mm) and no larger than 12 points (6mm) after reduction and must be consistent.

% Figures and tables should be labelled and referenced in the standard way using the \verb|\label{}| and \verb|\ref{}| commands.

% Fig. \ref{fig:frog} shows an example of how to insert a column-wide figure. To insert a Fig. wider than one column, please use the \verb|\begin{figure*}...\end{figure*}| environment. Figures wider than one column should be sized to 11.4 cm or 17.8 cm wide. Use \verb|\begin{SCfigure*}...\end{SCfigure*}| for a wide Fig. with side legends.

% \subsection*{Tables}
% Tables should be included in the main manuscript file and should not be uploaded separately.

% \begin{table}%[tbhp]
% \centering
% \caption{Comparison of the fitted potential energy surfaces and ab initio benchmark electronic energy calculations}
% \begin{tabular}{lrrr}
% Species & CBS & CV & G3 \\
% \midrule
% 1. Acetaldehyde & 0.0 & 0.0 & 0.0 \\
% 2. Vinyl alcohol & 9.1 & 9.6 & 13.5 \\
% 3. Hydroxyethylidene & 50.8 & 51.2 & 54.0\\
% \bottomrule
% \end{tabular}

% \addtabletext{nomenclature for the TSs refers to the numbered species in the table.}
% \end{table}

% \subsection*{Supporting Information Appendix (SI)}

% Authors should submit SI as a single separate SI Appendix PDF file, combining all text, figures, tables, movie legends, and SI references. SI will be published as provided by the authors; it will not be edited or composed. Additional details can be found in the \href{https://www.pnas.org/authors/submitting-your-manuscript#manuscript-formatting-guidelines}{PNAS Author Center}. The PNAS Overleaf SI template can be found \href{https://www.overleaf.com/latex/templates/pnas-template-for-supplementary-information/wqfsfqwyjtsd}{here}. Refer to the SI Appendix in the manuscript at an appropriate point in the text. Number supporting figures and tables starting with S1, S2, etc.

% Authors who place detailed materials and methods in an SI Appendix must provide sufficient detail in the main text methods to enable a reader to follow the logic of the procedures and results and also must reference the SI methods. If a paper is fundamentally a study of a new method or technique, then the methods must be described completely in the main text.

% \subsubsection*{SI Datasets}

% Supply .xlsx, .csv, .txt, .rtf, or .pdf files. This file type will be published in raw format and will not be edited or composed.

%EXTRA TEXT

% Without an inner core, Earth's magnetic field would have been generated solely through CMB heat flow, with perhaps additional stirring by other power sources such as Mg precipitation \cite{ORourke2016a}, or by exotic processes such as a convecting basal magma ocean \cite{Ziegler2013a}. Long-term secular cooling of Earth has the potential to have decreased the power available to the geodynamo through time prior to the nucleation of the inner core \cite[e.g][]{Davies2021a}. As a result, it is possible that intensity of the geomagnetic axial dipole field would have experienced a long-term decay before the emergence of the solid inner core after which it would have strengthened \cite{Aubert2009a, Davies2021a}. However, the estimates for thermal conductivity values for the core continue to vary \cite{Konopkova2016a, Ohta2016a, Zhang2020a}. Obtaining an observational record of the timing of inner core formation would constrain the thermal evolution of the core and mantle, in addition to improving our understanding of the history of Earth’s dynamo.

% Interpretations of selected records of the current Precambrian absolute paleointensity database (compiled in the PINT database; \url{http://earth.liv.ac.uk/pint/}) suggest that an appreciable geomagnetic field existed in the Archean but reached its weakest in the Ediacaran when very low virtual dipole moment (VDM) estimates have been obtained \cite{Bono2019a, Shcherbakova2019a, Thallner2021b, Lloyd2021a}, followed by a recovery near the Ediacaran-Cambrian boundary \cite{Thallner2021a}. Ref. \citealp{Bono2019a} adopted a billion-year-scale view of such observations to suggest that these data are consistent with thermal evolution models that use high core thermal conductivity values, where the nucleation of the inner core did not begin to provide compositional convection power for driving the geodynamo until the Ediacaran-Cambrian, suggesting that the Proterozoic dynamo might have experienced a monotonic decay when compositional convection was lacking and thermal convection was progressively decreasing \cite{Driscoll2016a, Davies2021a}.

% Other studies focusing on paleomagnetic directions have also found that the late Ediacaran might have been a period of time when the geomagnetic field were experiencing frequent reversals and waning of dipole dominance \cite{Bono2015a, Kodama2020a}.

% However, previous paleointensity data from intrusive and extrusive rocks in the ca. 1.1 Ga Midcontinent Rift \cite{Pesonen1983a, Kulakov2013a, Sprain2018a} record VDM values similar to that of today, which are higher than the values predicted by the long term decay curve proposed by  \citealp{Bono2019a} (Fig. \ref{fig:PINT_compilation}). Ref. \citealp{Pesonen1983a} obtained an average virtual dipole moment of 93 $\times 10^{21}$ Am$^2$ (ZAm$^2$) after applying a 13\% cooling rate correction (114 ZAm$^2$ before correction) using both extrusive and intrusive rocks from the Midcontinent Rift, including the ca. 1108-1105 Ma Osler Volcanics, the ca. 1100 Ma Maimainse Point volcanics, the ca. 1107 Ma intrusive Logan Sills, Thunder Bay dikes, the Baraga and Marquette dikes, as well as lava flows of the ca. 1096 Ma North Shore Volcanic Group (Fig. \ref{fig:Geologic_map}). Ref. \citealp{Kulakov2013a} obtained an average VDM of 59 $\pm$ 11 ZAm$^2$ using the ca. 1085 Ma Lake Shore Traps volcanics and suggest that the present-day-like geomagnetic field strength likely existed through the late stage of Midcontinent Rift magmatism \cite{Vervoort2007a, Miller2013a}. However, non-ideal paleointensity behaviors at specimen level and overall low success rates have challenged the interpretation of many previous paleointensity results obtained by  \citealp{Pesonen1983a, Kulakov2013a}. One typical failure of paleointensity experiment is the double-slope behavior, as two distinct paleointensity estimates may be calculated depending on the interpreter's choice of slope: a higher paleointensity estimate from the low-temperature portion and a lower paleointensity estimate from the high-temperature portion. Ref \citealp{Pesonen1983a} used the low-temperature slope as the best representation of the past magnetic field strength (likely overestimating the field strength) whereas  \citealp{Kulakov2013a} used the high-temperature slope (likely underestimating the field strength). Results with such non-ideal behavior were rejected by  \citealp{Sprain2018a} who applied stricter paleointensity selection criteria on new paleointensity results and accepted only 23 out of 133 specimens (5 out of 21 lava flows) from the Osler Volcanics and the Mamainse Point volcanics. A calculated mean VDM value of 56 $\pm$ 21 ZAm$^2$ is similar to the average value of the past 300 million years \cite{Sprain2018a}.

% Although it is typical of paleointensity experiments to have low success rates, bulk plagioclase-rich rocks \cite{Selkin2000a} and single plagioclase crystals \cite{Tarduno2005a} are found to be able to produce high quality paleointensity results. It has been suggested that plagioclase crystals can protect magnetic minerals that exsolved within them against post-formation weathering and thermochemical alteration during paleointensity experiments \cite{Tarduno2005a}. In this study, we present new paleointensity results from IZZ modified absolute paleointensity experiments \cite{Yu2004a} on a suite of bulk samples from ca. 1092 Ma anorthosite xenoliths and their Beaver River diabase host rocks of the Beaver Bay Complex within the Midcontinent Rift. Paleointensity experiments on the anorthosite xenoliths have a very high success rate. Their consistent specimen- and site-level paleointensity results together with rock magnetic data reveal that the anorthosite xenoliths have low anisotropy of thermal remanent magnetization (TRM) and can acquire TRM linearly. Magnetic imaging shows the anorthosite specimens have dominant magnetic carriers interstitial to and within plagioclase crystals and they do not display strong preferred orientations. Our single-slope, high quality paleointensity data add to the reliable paleointensity records during the $\sim$25 Myr of Midcontinent Rift magmatic activity. Our new data from the Midcontinent Rift intrusive rocks support previous data from the volcanics that there was a strong late Mesoproterozoic geodynamo.

>>>>

The lack of symplecites and the lack of large plagioclase-hosted titanomagnetite needles within the Beaver River anorthosite xenoliths targeted in this study distinguish them from plagioclase cumulates of the Duluth Complex and other layered mafic complexes where anisotropic magnetic mineral fabrics associated with igneous foliation often occur \cite{Scofield1986a, Selkin2000a, Feinberg2006a}. The relative lack of fabric and associated preferred orientation makes the Beaver River anorthosite xenoliths a particularly compelling target for paleointensity experiments.

@@

In addition to estimating paleointensity values by introducing a set of selection criteria to filter our experiment results, we apply an independent method of \citealp{Cych2021a} which uses all experimental data regardless of their NRM/TRM plot statistics to perform bias corrected estimation of paleointensity (BiCEP). This Bayesian probabilistic method is based on an assumption that paleointensity estimates from specimens that come from a same cooling unit are distributed around a true paleointensity value with the various deflections being expressed as the curvature parameter of the NRM/TRM plot \cite{Paterson2011a}. The posterior paleointensity distributions from these sites with high-quality specimen-level data are in agreement with the site-level averages developed using the selection criteria approach (Fig. \ref{fig:PINT_cooling_corrected}; Fig. S1). Overall, the high-quality results from the anorthosite xenoliths of the Beaver River diabase indicate that the anorthosite xenoliths record a high geomagnetic field ca. 1092 Ma.

@@

The hypothesis that a weak Ediacaran geomagnetic field is a telltale sign of the lack of an inner core with core nucleation following shortly thereafter may predict that it is the most significant weak to strong field transition in the paleointensity record. However, Fig. \ref{fig:PINT_compilation} shows that transitions from low to high field intensities occurred before, during, and after the Ediacaran. In the Ediacaran record developed to date, there is a two-fold increase in Earth's virtual dipole moment when comparing estimates from the ca. 565 Ma Sept-$\hat{I}$les intrusions \cite{Bono2019a} to those from ca. 550 Ma volcanics of the Skinner Cove Formation \cite{Thallner2021a} (Fig. \ref{fig:PINT_compilation}). In the late Mesoproterozoic, there is at least a six-fold increase within a period of $\sim$35 Myr from a low average virtual dipole moment of $\sim$13 ZAm$^2$ recorded by the ca. 1140 Ma Abitibi dikes which yielded straight Arai plots \cite{Macouin2003a}, to a high moment of $\sim$70 ZAm$^2$ recorded by the ca. 1106 Ma Osler Volcanics, with even stronger values from ca. 1092 Ma by Beaver River anorthosite xenoliths that record virtual dipole moments up to $\sim$140 ZAm$^2$. While the Abitibi dikes paleointensity estimates do not go to values as low as the Sept-$\hat{I}$les intrusions, this virtual dipole moment increase in the Mesoproterozoic is the largest yet documented in the Precambrian on a 10s of millions of years timescale (Fig. \ref{fig:PINT_compilation}). The tempo and scale of this field intensity transition could match with model-based predictions associated with the onset of inner core nucleation \cite{Davies2021a}. This timing would be broadly consistent with that proposed by \citealp{Biggin2015a}. However, a model prediction could be of sustained strong field values following inner core nucleation which is challenged by data from the ca. 1070 Ma Bangemall Sills which include a sill with a low virtual dipole moment of $\sim$6.4 ZAm$^2$ (Fig. \ref{fig:PINT_compilation};  \citealp{Lloyd2021b}). Following the Ediacaran, there are also low paleointensity estimates from Devonian rocks such as ca. 370 Ma dikes and lavas of the Siberian Viluy Traps that give virtual dipole moment estimates of 4.3 to 14.9 ZAm$^2$ (Fig. \ref{fig:PINT_compilation}; \citealp{Hawkins2019a}). These low values as well as data from the ca. 414 Ma Strathmore lava flows \cite{Hawkins2021a}, the ca. 410-380 Ma lava flows of Siberia and the Kola Peninsula \cite{Shcherbakova2017a}, the ca. 408–393 Ma Buribay volcanics \cite{Shcherbakova2021a} and the ca. 332 Ma Kinghorn volcanics \cite{Hawkins2021a} has led to the proposal of this interval as the ``Mid-Paleozoic Dipole Low'' \cite{Hawkins2021a}. This ``Mid-Paleozoic Dipole Low'' is followed by high paleointensity values such that there is a six-fold increase in the database from virtual dipole moment of $\sim$16 ZAm$^2$ at ca. 332 Ma to $\sim$99 ZAm$^2$ ca. 308 Ma in the late Carboniferous (Fig. \ref{fig:PINT_compilation}; \citealp{Hawkins2021a}).

Given that the multiple records of a weak field can not all be a minimum prior to inner core nucleation, what processes could lead to a weak dipole at Earth's surface even in the presence of a crystallizing inner core? Numerical models have shown that the dipole moment is sensitive to both the magnitude and spatial pattern of heat flow across the core-mantle boundary when there are strong available power sources to the geodynamo \cite{Olson2007a, Olson2010a}. In such models, relatively low total heat flux across the core-mantle boundary can prevent the axial dipole from reversing whereas a high heat flux through the boundary can result in an increase in reversal frequency and decrease in dipole intensity. The ``Mid-Paleozoic Dipole Low'' has been hypothesized to be the result of such elevated core-mantle boundary heat flux conditions at a time when there was also available power from a crystallizing inner core \cite{Hawkins2019a}, thereby explaining the observed low paleointensities which include values as weak as those of the ca. 565 Ma Sept-$\hat{I}$les intrusions \cite{Bono2019a} (Fig. \ref{fig:PINT_compilation}). Mantle convection can modulate core mantle boundary heat flow through changes in the structure of the deep mantle associated with upwelling plumes \cite{Larson1991a, Courtillot2007a} and subducted slabs \cite{Tan2002b, Biggin2012a, Hounslow2018a}. Strong evidence for differential plate tectonic motion extends back to ca. 2.2 Ga in the Paleoproterozoic \cite{Mitchell2014a,Swanson-Hysell2021b} and potentially back to ca. 3.2 Ga in the Archean \cite{Brenner2020a}. Plate tectonic modulations of core mantle boundary heat flow are therefore expected throughout the Proterozoic. Such changes may explain large variability in Proterozoic paleointensity values similar to those seen in the Phanerozoic \cite{Lloyd2021a} and may challenge our ability to detect the increase in surface geomagnetic field strength predicted to have happened at the onset of inner core crystallization.

Overall, the high-fidelity paleointensity recorders of the Beaver River anorthosite xenoliths in the well-preserved Midcontinent Rift record strong field strengths 1.1 billion years ago. The highest site-level value of the virtual dipole moment would rank in the top 1$\%$ of those in the database for the Cenozoic Era (the last 66 million years) when there was unequivocally a crystallizing inner core. These high surface field strengths necessitate appreciable power to the late Mesoproterozoic geodynamo.

@@

@@

@@
